Improving Item-Item Similarity Estimation during Collaborative Filtering

### 2014-03-25 paper/paper.tex
Fix latex error (\hat only allowed in math mode).
Change section titles for Model-based Approach.
Develop Probabilistic Model subsection of Model-based Approach.

The problem of approximating the $TrueSim$ from $Sim_n$ can now be seen as
finding the value of $Y$ that most likely produced $X_n$. In other words we 
desire the Maximum Likelihood Estimate of $Y | X_n$:
\begin{align}
\hat{y} &= \argmax_yP(Y=y|X_n=x) 
\\&= \argmax_y\left[\frac{P(X_n=x | Y=y)P(Y=y)}{P(X_n=x)}\right]
\\&= \argmax_y\left[P(X_n=x|Y=y)P(Y=y)\right]
\\&= 
\argmax_y\left[\frac{1}{\sigma_{2,n}\sqrt{2\pi}}\exp{\left(\frac{-(x-y)^2}
{2\sigma_{2,n}^2}\right)}
\frac{1}{\sigma_{1}\sqrt{2\pi}}\exp{\left(\frac{-(y-\mu)^2}
{2\sigma_{1}^2}\right)}\right]
\\&= \argmin_y\left[\frac{(x-y)^2}{2\sigma_{2,n}^2} +
\frac{(y-\mu)^2}{2\sigma_{1}^2}\right]
\\&= \argmin_y\left[\left(\sigma_{1}^2+\sigma_{2,n}^2\right)y^2 - 
2\left(\sigma_{1}^2x+\sigma_{2,n}^2\mu\right)y\right]
\end{align}
We find the exact minimum by taking the derivative with respect to $y$ and
setting it to zero:
\begin{align}
\frac{d}{dy}\left[\left(\sigma_{1}^2+\sigma_{2,n}^2\right)y^2 - 
2\left(\sigma_{1}^2x+\sigma_{2,n}^2\mu\right)y\right]
&= 2\left(\sigma_{1}^2+\sigma_{2,n}^2\right)y - 
2\left(\sigma_{1}^2x+\sigma_{2,n}^2\mu\right) 
\\&= 0
\end{align}
This produces the following linear equation for \hat{y} in terms of x:
\begin{align}
\hat{y} &= \frac{\sigma_{1}^2x+\sigma_{2,n}^2\mu}{\sigma_{1}^2+\sigma_{2,n}^2}
\end{align}
which can be rewritten as:
\begin{align}
\left(\hat{y} - \mu\right) &= \frac{\sigma_{1}^2}{\sigma_{1}^2+\sigma_{2,n}^2}
\left(x-\mu\right)
\end{align}

@@

\section*{A Model-based Technique}

@@

Let $Y$ be a random variable representing the $TrueSimRand$ of a randomly chosen
pair of items. As before, we model $Y$ as a Normal distribution
\begin{align}
Y \sim N(\mu_s, \sigma_s^2),
\end{align}
where $\mu_s$ and $\sigma_s$ are the average and standard deviation of all
similarity scores, respectively. Now, let $\vec{R} = \left(R_A~~R_B\right)$ be
a random two element vector representing the ratings given to items $A$ and $B$.
We model the conditional distribution of $\vec{R} | Y$ as a multivariate Normal
distribution
\begin{align}
\vec{R} | Y=y \sim N(\vec{\mu_r}, \mathbf{\Sigma_y}),
\end{align}
where the elements of $\vec{\mu_r}$ are the average ratings and
$\mathbf{\Sigma_y}$ is the $y$-dependent $2\times2$ covariance matrix.

>>>>

The problem of approximating the $TrueSim$ from $Sim_n$ can now be seen as
finding the value of $Y$ that most likely produced $X_n$. In other words we 
desire the Maximum Likelihood Estimate of $Y | X_n$:
\begin{align}
\hat{y} &= \argmax_yP(Y=y|X_n=x) 
\\&= \argmax_y\left[\frac{P(X_n=x | Y=y)P(Y=y)}{P(X_n=x)}\right]
\\&= \argmax_y\left[P(X_n=x|Y=y)P(Y=y)\right]
\\&= 
\argmax_y\left[\frac{1}{\sigma_{2,n}\sqrt{2\pi}}\exp{\left(\frac{-(x-y)^2}
{2\sigma_{2,n}^2}\right)}
\frac{1}{\sigma_{1}\sqrt{2\pi}}\exp{\left(\frac{-(y-\mu)^2}
{2\sigma_{1}^2}\right)}\right]
\\&= \argmin_y\left[\frac{(x-y)^2}{2\sigma_{2,n}^2} +
\frac{(y-\mu)^2}{2\sigma_{1}^2}\right]
\\&= \argmin_y\left[\left(\sigma_{1}^2+\sigma_{2,n}^2\right)y^2 - 
2\left(\sigma_{1}^2x+\sigma_{2,n}^2\mu\right)y\right]
\end{align}
We find the exact minimum by taking the derivative with respect to $y$ and
setting it to zero:
\begin{align}
\frac{d}{dy}\left[\left(\sigma_{1}^2+\sigma_{2,n}^2\right)y^2 - 
2\left(\sigma_{1}^2x+\sigma_{2,n}^2\mu\right)y\right]
&= 2\left(\sigma_{1}^2+\sigma_{2,n}^2\right)y - 
2\left(\sigma_{1}^2x+\sigma_{2,n}^2\mu\right) 
\\&= 0
\end{align}
This produces the following linear equation for $\hat{y}$ in terms of $x$:
\begin{align}
\hat{y} &= \frac{\sigma_{1}^2x+\sigma_{2,n}^2\mu}{\sigma_{1}^2+\sigma_{2,n}^2}
\end{align}
which can be rewritten as:
\begin{align}
\left(\hat{y} - \mu\right) &= \frac{\sigma_{1}^2}{\sigma_{1}^2+\sigma_{2,n}^2}
\left(x-\mu\right)
\end{align}

@@

\section*{Model-based Approach}
\subsection*{Probabilistic Model}

@@

Let $Y$ be a random variable representing the $TrueSimRand$ of a randomly chosen
pair of items. As before, we model $Y$ as a Normal distribution
\begin{align}
Y \sim N(\mu_s, \sigma_s^2),
\end{align}
where $\mu_s$ and $\sigma_s^2$ are the average and variance of all similarity
scores, respectively. Now, let $\vec{R} = \left(R_A~~R_B\right)$ be a random two
element vector representing the ratings given to items $A$ and $B$.  We model
the conditional distribution of $\vec{R} | Y$ as a multivariate Normal
distribution
\begin{align}
\vec{R} | Y=y \sim N(\vec{\mu_r}, \mathbf{\Sigma_y}),
\end{align}
where the elements of $\vec{\mu_r}$ are the average ratings and
$\mathbf{\Sigma_y}$ is the $y$-dependent $2\times2$ covariance matrix. Now,
being that $Y=y$ is the Pearson correlation of the ratings $R_A | y$ and $R_B |
y$ generated by this distribution, if we assume the variance of individual
ratings $\sigma_r^2$ is independent of $y$, we have that
\begin{align}
y = \frac{\mathrm{COV}(R_A | y, R_B | y)}{\sigma_r^2},
\end{align}
where $\mathrm{COV}(R_A | y, R_B | y)$ is the covariance of the ratings $R_A |
y$ and $R_B | y$.  It therefore follows that if the ratings are indeed normally
distributed that the covariance matrix is related to the similarity $y$ by
\begin{align}
\mathbf{\Sigma_y} = \sigma_r^2 \cdot
\left[ \begin{array}{cc}  1 & y \\ y & 1 \end{array} \right]
\end{align}
